Moisture effects on leaf litter are species-specific and result in non-additive flammability in mixed conifer forest

### 2019-08-05 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Include Rita's comments 2019-08-05

      \item Moisture content is a strong determinant of forest fuel
    flammability and varies on small spatial scales and over short time
    periods. Empirical modeling of equilibrium dead fine fuel moisture is well
    established, but less is known about the mechanisms by which
    species-specific litter traits influence litter moisture dynamics.
    Characterizing how species’ leaf litter retain and release moisture through
    time is critical to the assessment of fuel’s availability to burn.
      \item Natural forests are often comprised of mixed stands, leading to
    mixtures of leaf litter on the surface fuelling the fires. Multiple studies
    have established the existence of non-additive effects in the flammability
    of dry litter, but we lack an understanding of moisture synergies in mixed
    litter and of moisture effects on the flammability of mixed litter.
      \item The litter from eight species of a mixed-conifer forest was
    saturated and allowed to dry to determine moisture absorption capacity and
    drying rates. Burn trials were performed in moist litter beds of single
    species and of mixtures to establish flammability response to the dry-down
    process.
      \item Moisture dynamics varied across species with higher specific leaf
    area being associated with higher maximum absorption and lower litter bulk
    density being associated with faster drying rates. Species that produce
    more aerated litter beds (such as \emph{Q. kelloggii} and \emph{Pinus}
    species) have faster rates of litter drying.
      \item We examined two axes of flammability variation (spread rate and
    fuel consumption) and found that these differing moisture dynamics result
    in time since wetting having strongly differing effects on flammability
    across the major litter types typified by different tree taxa.
      \item We found that litter mixtures exhibited consistent non-additivity
    in flame spread rate in fuel consumption. This non-additivity became more
    positive with fuel drying. Litter mixtures burn with fire behavior more
    similar to that of the most flammable constituent species and this effect
    increases as fuels dry.
      \item \emph{Synthesis:} In this temperate mixed conifer forest, the shade
    intolerant species are favored by the increased fire to which their fuels
    contribute Such positive feedback effects increases the potential for
    species to exert community scale effects even when relatively rare and that
    can influence community assembly and the evolutionary of niche-constructing
    traits.
  \end{enumerate}
\end{abstract}

@@

Litter packing (measured as litter-bed density or packing ratio) controls the
flammability of dry litter fuels in seasonally dry climates because higher
litter density reduces flammability, especially on the flame spread rate axis.
Leaf litter density is driven by leaf size and shape
\citep{Fonda+Belanger+etal-1998, Scarff+Westoby-2006, Kane+Varner+etal-2008}.
These litter particle effects on litter bed aeration, in turn, have effects on
litter bed flammability \citep{Ganteaume+Marielle+etal-2011,
  Schwilk+Caprio-2011, Kreye+Varner+etal-2013}. These species-specific effects
are well established. However, how moisture dynamics modulate species specific
flammability is poorly known, and could be driven by by leaf traits, emergent
litter litter traits, or both. Species specific effects on litter flammability
are recorded for the forests of the southeastern US \citep{Nowacki+Abrams-2008}
where species-specific litter drying rates play a role in flammability
feedbacks \citep{Kreye+Varner+etal-2013}. Similarly, in mixed-conifer forest of
the western USA, where the fire regime is characterized by surface fires fueled
by plant litter, overstory tree composition drives variation in local fire
behavior \citep{Schwilk+Caprio-2011} as a result of differential litter
flammability across tree species and as a result of synergistic interactions in
multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}. These flammability
feedbacks may help explain historical fire frequency patterns, but current
estimates of litter flammability in mixed conifer forests are based largely on
burning dry litter.

@@

The overall ranking of flammability in litter mixtures showed some consistency
with the results obtained for the individual species, but there were
significant non-additive effects. Mixtures containing \emph{Q. kelloggii} and
\emph{P. jeffreyi} tended to have fast spread rates. These results are likely
influenced by the lower bulk densities of litter mixtures containing those two
species. Most interestingly, there were striking pattern of non-additivity in
spread rate and in fuel consumption for the four mixtures, indicating that
positive non-additivity in flammability is not restricted to completely dry
litter mixtures observed in previous work \citep{Magalhaes+Schwilk-2012,
  VanAltena+Logtestjin+etal-2012}. Non-additivity in both spread rate and fuel
consumption became more positive with time since wetting and these patterns
were consistent across mixture identities. This is a slightly different result
than that reported by Blauw et al \citeyear{Blauw+Wensink+etal-2015}. Blauw et
al found that non-additivity was greater in moist mixtures of heathland fuels
than in dry fuels. In their work, non-additivity was generally negative but cases of 
positive non-additivity in mixtures was often a result of mixtures meeting
ignition thresholds not met by the constituent species alone. To limit the
number of models fit, we limited our investigation to two flammability
parameters that represent two major axes of flammability variation. Therefore,
our study does not fully capture ignition effects although we do have some
results consistent with theirs: mixtures with \emph{Q. kellogii} could ignite
at times since wetting at which the species in monoculture could not.

@@

The authors thank the National Park Service and the employees
  of Sequoia and Kings Canyon National Parks.This work was carried out with
  permission from NPS under permit \# SEKI-2014-SCI-0040. We also thank
  students who assisted with field work and the litter collection, including
  Lizz Waring, Taylor Brown and Chelsea Griffin. We thank the Lubbock Fire
  Department for their ongoing help and cooperation.

>>>>

      \item Moisture content is a strong determinant of forest fuel
    flammability and varies on small spatial scales and over short time
    periods. Empirical modeling of equilibrium dead fine fuel moisture is well
    established, but less is known about the mechanisms by which
    species-specific litter traits influence litter moisture dynamics.
    Characterizing how species’ leaf litter retain and release moisture through
    time is critical to the assessment of fuel’s availability to burn.
      \item Natural forests are often comprised of mixed stands, leading to
    mixtures of leaf litter on the surface fuelling the fires. Multiple studies
    have established the existence of non-additive effects in the flammability
    of dry litter, but we lack an understanding of moisture synergies in mixed
    litter and of moisture effects on the flammability of mixed litter.
      \item The litter from eight species of a mixed-conifer forest was
    saturated and allowed to dry to determine moisture absorption capacity and
    drying rates. Burn trials were performed in moist litter beds of single
    species and of mixtures to establish flammability response to the dry-down
    process.
      \item Moisture dynamics varied across species with higher specific leaf
    area being associated with higher maximum absorption and lower litter bulk
    density being associated with faster drying rates. Species that produce
    more aerated litter beds (such as \emph{Q. kelloggii} and \emph{Pinus}
    species) have faster rates of litter drying.
      \item We examined two axes of flammability variation (spread rate and
    fuel consumption) and found that these differing moisture dynamics result
    in time since wetting having strongly differing effects on flammability
    across the major litter types typified by different tree taxa.
      \item We found that litter mixtures exhibited consistent non-additivity
    in flame spread rate in fuel consumption. This non-additivity became more
    positive with fuel drying. Litter mixtures burn with fire behavior more
    similar to that of the most flammable constituent species and this effect
    increases as fuels dry.
      \item \emph{Synthesis:} In this temperate mixed conifer forest, the shade
    intolerant species are favored by the increased fire to which their fuels
    contribute Such positive feedback effects increases the potential for
    species to exert community scale effects even when relatively rare and that
    can influence community assembly and the evolution of niche-constructing
    traits.
  \end{enumerate}
\end{abstract}

@@

Litter packing (measured as litter-bed density or packing ratio) controls the
flammability of dry litter fuels in seasonally dry climates because higher
litter density reduces flammability. Leaf litter density is driven by leaf size
and shape \citep{Fonda+Belanger+etal-1998, Scarff+Westoby-2006,
  Kane+Varner+etal-2008}. These litter particle effects on litter bed aeration,
in turn, have effects on litter bed flammability
\citep{Ganteaume+Marielle+etal-2011, Schwilk+Caprio-2011,
  Kreye+Varner+etal-2013}. These species-specific effects are well established.
However, how moisture dynamics modulate species specific flammability is poorly
known, and could be driven by by leaf traits, emergent litter litter traits, or
both. Species specific effects on litter flammability are recorded for the
forests of the southeastern US \citep{Nowacki+Abrams-2008} where
species-specific litter drying rates play a role in flammability feedbacks
\citep{Kreye+Varner+etal-2013}. Similarly, in mixed-conifer forest of the
western USA, where the fire regime is characterized by surface fires fueled by
plant litter, overstory tree composition drives variation in local fire
behavior \citep{Schwilk+Caprio-2011} as a result of differential litter
flammability across tree species and as a result of synergistic interactions in
multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}. These
flammability feedbacks may help explain historical fire frequency patterns, but
current estimates of litter flammability in mixed conifer forests are based
largely on burning dry litter.

@@

The overall ranking of flammability in litter mixtures showed some consistency
with the results obtained for the individual species, but there were
significant non-additive effects. Mixtures containing \emph{Q. kelloggii} and
\emph{P. jeffreyi} tended to have fast spread rates. These results are likely
influenced by the lower bulk densities of litter mixtures containing those two
species. Most interestingly, there were striking patterns of non-additivity in
spread rate and in fuel consumption for the four mixtures, indicating that
positive non-additivity in flammability is not restricted to completely dry
litter mixtures observed in previous work \citep{Magalhaes+Schwilk-2012,
  VanAltena+Logtestjin+etal-2012}. Non-additivity in both spread rate and fuel
consumption became more positive with time since wetting and these patterns
were consistent across mixture identities. This is a slightly different result
than that reported by Blauw et al \citeyear{Blauw+Wensink+etal-2015}. Blauw et
al found that non-additivity was greater in moist mixtures of heathland fuels
than in dry fuels. In their work, non-additivity was generally negative but cases of 
positive non-additivity in mixtures was often a result of mixtures meeting
ignition thresholds not met by the constituent species alone. To limit the
number of models fit, we limited our investigation to two flammability
parameters that represent two major axes of flammability variation. Therefore,
our study does not fully capture ignition effects although we do have some
results consistent with theirs: mixtures with \emph{Q. kellogii} could ignite
at times since wetting at which the species in monoculture could not.

@@

The authors thank the National Park Service and the employees of Sequoia and
Kings Canyon National Parks. This work was carried out with permission from NPS
under permit \# SEKI-2014-SCI-0040. The authors thank Russell Lackey who
contributed essential assistance with the field work and Gabrielle Plata and
Leticia Plata who assisted with the burning trials. We thank the Lubbock Fire
Department for their ongoing help and cooperation.
